Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm

### 2014-10-22 paint-contour-tree.tex
section 6 proofs

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{high}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.

@@

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
For $x$, $\col(x)$ is the set of colors that touch $x$.
\end{definition}
%
%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

\medskip
\noindent
{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
heap $T(c)$ of non-extremal critical points, keyed by vertex heights.

@@

\medskip
\noindent
{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
are in these sets are all constant time operations. The stack is guaranteed to satisfy 
the following invariants.
\begin{asparaitem}
	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
	in $T(c)$. Furthermore, $c = \rep(c)$.
	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
\end{asparaitem}

\medskip
\noindent
{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
with the algorithm.

@@

We have a simple, but critical, procedure that updates the stack $\st$ by pushing
points on to it.

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and add $c'$ to $\mcol(x)$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

\medskip
When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
$h$ is the highest point in $\touch(c)$.

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
vertices.

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
\begin{asparaitem}
	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.

{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
sets of colors with vertices.

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
\end{definition}

Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.

\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.

@@

%
%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

\medskip
\noindent
{\bf The binomial heaps $\touch(c)$:} For each color $c$, $T(c) = \{v | c \in \fcol(v)\}$.
This is stored as a \emph{binomial max-heap} $T(c)$, keyed by vertex heights.

@@

\medskip
\noindent
{\bf The stack $\st$:} This consists of non-extremal critical points, with monotonically increasing
heights as we go from the base to the head.
\ignore{
Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
are in these sets are all constant time operations. The stack is guaranteed to satisfy 
the following invariants.
\begin{asparaitem}
	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
	in $T(c)$. Furthermore, $c = \rep(c)$.
	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
\end{asparaitem}
}

\medskip
\noindent
{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.

\subsection{Properties of a painted simplicial complex} \label{sec:prop}

\begin{lemma} \label{lem:full} Consider vertex $v$ and take $(f(v)+\delta)$-contour $\phi$ intersecting
$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

\begin{proof} Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.
\end{proof}

\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \fcol(v)$, $v$
is the highest vertex in $T(c)$. Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(c)$ is empty.
\end{definition}

@@

We have an imporant procedure that updates the stack $\st$.

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
	\item Let $h$ be head of $\st$.
	\item While $h$ is not ripe:
	\begin{compactenum}
		\item (Either (i) there is $c \in \fcol(v)$ such that $v$ is not the highest in $T(c)$ or (ii) there is 
		$c \in \col(v) \setminus \fcol(v)$, such that $T(c)$ is non-empty.)
		\item In either case, push the highest of $T(c)$ onto $\st$, and update head $h$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

\medskip
Clearly, when $\update(\st)$ terminates, the head $h$ is ripe. The following crucial claim demonstrates
the utility of the stack.

\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
\end{claim}

\begin{proof} Consider a single push onto the stack. In Case (i), the increase in height is obvious.
Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.

All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
\end{proof}

The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
Of course, it is not possible to explicitly redo the painting, but this is implictly done
through the union-find data structure on colors. So, when we process some vertex $v$,
which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
have merged.
